---
title: "Introduction"
---

```{r eval = TRUE,  message=F, include=F, warning=F, purl=F, results="hide"}
knitr::purl("introduction.Rmd", documentation = F)
```

```{r echo=FALSE, purl=F}
xfun::embed_file('introduction.Rmd')
```

```{r echo=FALSE, purl=F}
xfun::embed_file('introduction.R')
```

```{r echo=FALSE, purl=F}
xfun::embed_file('Data/reef_fish.xlsx')
```



# R software

## What is R?

R is a popular free and open program for statistical computing (https://www.r-project.org/about.html). R is a widely-used program for **statistical computing**. R can be run from the command line or from a **graphical user interface (GUI)** that is automatically installed when you install R (if you open the R program installed on your computer, this is what will open up). 

R consists of a **core** (minimal installation + add on **packages** (from the Comprehensive R Archive Network [CRAN], Bioconductor, or others). The core's capabilities is expanded by using the add-on packages.

## Installation

### Windows

If you don't have R installed on your computer:

- Download R from the [CRAN website](https://cran.r-project.org/bin/windows/base/release.htm).
- Run the `.exe` file that was just downloaded.
- Once it’s installed, open R to make sure it works and you don’t get any error messages.

If you already have R already install:

To check the version of R you are using, start R: the version of R you are running is the first thing that appears on the terminal. Alternatively, Type `sessionInfo()` in the R console, which will  display which version of R you are running. Go on the [CRAN website](https://cran.r-project.org/) and check whether a more recent version is available. If so, please download and install it. You can check [here](https://cran.r-project.org/bin/windows/base/rw-FAQ.html#How-do-I-UNinstall-R_003f) for more information on how to remove old versions from your system if you wish to do so.

### macOS

If you don't have R installed on your computer:

- Download R from the [CRAN website](https://cran.r-project.org/bin/macosx/).
- Select the `.pkg` file for the latest R version.
- Double click on the downloaded file to install R
- It is also a good idea to install [XQuartz] (https://www.xquartz.org/) (needed by some packages)
- Once it’s installed, open R to make sure it works and you don’t get any error messages.

If you already have R already install:

To check the version of R you are using, start R: the version of R you are running is the first thing that appears on the terminal. Alternatively, Type `sessionInfo()` in the R console, which will  display which version of R you are running. Go on the [CRAN website](https://cran.r-project.org/bin/macosx/) and check whether a more recent version is available. If so, please download and install it. 


### Linux

- Follow the instructions for your distribution from [CRAN](https://cloud.r-project.org/bin/linux/), they provide information to get the most recent version of R for common distributions. 

- Once it’s installed, open R to make sure it works and you don’t get any error messages.

**NOTE**: the entire course is prepared using R for windows version. Despite most of the code provided will work on any versions, slight differences may cause trouble when we will deal with animation and 3D visualizations. 

## Update

If you already have R on your computer, an alternative is to run the following line of code in R native GUI (<span style="color: red">it will tell you if your R version is out-of-date </span>): 

```{r,  eval=FALSE, tidy=TRUE}
# R update (to run in R native GUI only)
if(!require(installr)) {
  install.packages("installr"); require(installr)} 
updateR()
```

This will start the updating process of your R installation.  It will check for newer versions, and if one is available, it will guide you through the decisions you need to make. 

You may have to choose a CRAN mirror if it is your first download. A mirror is just a copy of the CRAN archives. Choose one in Taiwan, it is usually a bit faster to download and install. 

## Customization

To expand the abilities of R, you can first install a package with its name. The package `ABC` provides extra tools (**functions**) for Approximate Bayesian Computation (ABC) 

- Install the package `abc`. To do only once.

```{r,  eval=FALSE}
# install the package `abc` with its name
install.packages("abc")
```

```{r,  eval=FALSE}
# function are case sensitive
# all information after `#` is ignored
# it is call a comment. Commenting is very important 
```

Happily, only basic packages are loaded when you started R. Other are inactive and you need to activate them. Once installed on your computer, you need to say that you want to use a specific package. 

- Use the package `abc`.To do every time you wanna use the package.

```{r,  eval=FALSE}
# use/load the package `abc` with its name
library("abc")
```

<p class="comment">
**Practice 1:** The `vegan` package provides tools for descriptive community ecology. It has most basic functions of diversity analysis, community ordination and dissimilarity analysis. Most of its multivariate tools can be used for other data types as well, not restricted to ecological purpose.Install and load the package `vegan`.
</p>

```{r, code_folding = 'Show Solution',  eval=FALSE}
# Install & load `vegan` package
install.packages('vegan')
library('vegan')
```

If you don't know the name of a package relevant for you analyses: 

- use task views available at http://cran.r-project.org/web/views. It group packages to subject areas such environmetrics, multivariate, etc. 

- use online discussion, forum, and docs. Among many others sources, you will often be directed to discussion in [stackoverflow](https://stackoverflow.com/)

## Help

You don't know: relax! You can access .html help via the 'Help' menu: *Help/Search help*. Alternatively, a faster solution is to call for help directly in the R console. You can get:

- a simple summary on how to use a specifically named function such as `mean`

```{r, code_folding = F, eval=T}
# simple help on the use of a function
median
```

- a detailed summary on the use of the function `median` which will send you to an .html webpage.

```{r, code_folding = F,  eval=FALSE}
# detailed help on the use of a function
?median # alternative 1
help(median) # alternative 2
```

- get a list of all CRAN functions with median in their description.

```{r,  eval=FALSE}
# using a keyword for search
??median
```

## Working directory

It is a good practice to create a working directory (folder) where you will store all your file. In general, each project will have its own working directory including data set, script, plot, etc. 

- retrieve your current working directory

```{r,  eval=F}
# get your current working directory
getwd()
```

You wanna change for a specific folder. Set it up using:

```{r,  eval=FALSE}
# set your current working directory
setwd()
```

## Calculator

R is a calculator:

```{r,  eval=FALSE}
# Use R as a calculator
3+2 # addition
3-2 # substraction
3*2 # multiplication
3/2 # division
3^3 # power
log(2) # logarithm
exp(2) # exponential
(5 + 3) / 4 # define priority using () or {} 
pi*4 # common function
```

R is case sensitive `pi` will work, whereas`Pi` will not 

## Objects

The list function `ls()` will provide you with all the objects in the memory.

```{r,  eval=T}
# my current list is empty 
ls()
```


```{r,  eval=T}
# I create and list 3 objects 
a<-'corals' # create an object 'a' containing 'corals'
b<-'are' # create an object 'b' containing 'are'
c<-'cool'# create an object 'c' containing 'cool'
ls() # list objects 'a', 'b', 'c'
```

**Note**: two operators `<-` and `=` can be used to import data. While longer, I favor `<-` in the creation of object to avoid confusion with mathematical operator

You can clean `rm` (remove) the objects `list`  in memory `ls()` using:

```{r,  eval=T}
# clean objects in memory
rm(list=ls())
```

**Tip**: The use of CTRL+L cleans your screen when you want to make it clearer. However, it **does not remove the objects** in memory.

## Reading

To be able to work with a dataset (and make calculation based on it), it has to be import as an object in the memory of R. Many formats can be used (.xls, .spss, SAS, etc.). To start with, we will use and .xlsx file, but quickly we will turn toward less 'heavy' format. Note the distinction between **reading** a file and **importing** a file.

My file `reef_fish` is an .xlsx file located in the file 'Data' of my working directory. 

```{r,  eval=T}
# 1. using the package `readxl`
# 2. reading my `reef_fish.xlsx` in my working directory
# 3. importing `reef_fish.xlsx` in a `fish` object
library(readxl) # load the package `readxl'
read_excel('Data/reef_fish.xlsx') # automatically print on my screen
fish<-read_excel('Data/reef_fish.xlsx') # store my table in an object called `fish`
fish # print my object `fish`   
```

More commonly, we will use .txt file (lighter) and the function `read.table'. But often you will need to think at several important **arguments** about the file you wanna to import:

- What is the exact name of the data file (with its extension)?
- Does the file have column titles?
- What is the delimiter between the columns (tabulation, space, comma or semicolon)?
- What is the character used for decimal points (dot, comma)?

```{r,  eval=T}
# importing a .txt file
fish<-read.table('Data/reef_fish.txt', header=T, sep='\t', dec='.') 
```

Check `?read.table` for all arguments you can specify within this function

It is not a good practice, but instead or working within  a working directory, you can also target directly your file (check your file properties) or an url.

```{r,  eval=FALSE,  tidy=TRUE}
# import file by path name
fish<-read.table ("D:/.../Topic 1/Data/reef_fish.txt",header = TRUE,sep="\t", dec=".")# long version
fish<-read.table("D:/.../Topic 1/Data/reef_fish.txt", TRUE, "\t",".")# short version
```

You can also use `file.choose()' to locate directly the file on computer

```{r,  eval=FALSE,  tidy=TRUE}
fish<-read.table(file.choose (), header = TRUE,sep="\t", dec=".")
```

**Note**: every time we created the object `fish` it replaced the one previously created.


<p class="comment">
**Practice 2:** Download `reef_fish.xlsx`, save it in a working directory you dedicate for this course, and import it in R. Do the same after converting this `reef_fish` file into a `.txt` file.
</p>

```{r, code_folding = 'Show Solution',  eval=FALSE}
# Too fast you need to try by yourself?
# Maybe a look at ?write.table may help
```

## Editor

You can start to feel it quickly becomes hard to manage the copy and paste + edit the function in the R GUI. It starts to be interesting to have an **script editor** to edit our code. It will help at saving and running efficiently our code in R. Many alternatives exist from opening a simple.txt file, opening an new script window directly in R, or using a dedicated script editor. The latest one is the most common alternative.

## Quit R

**File -> Exit** or simply type in the R console:

```{r,  eval=FALSE}
q()
```

Saving the workspace image will allow you to save all objects created in the R memory into your working directory (.Rdata). It also saves all commands given during the session (.Rhistory). Quit **without saving**.

# RStudio

## What's RStudio?

RStudio is a graphical integrated development environment (IDE) that makes using R much easier and more interactive. It is open source (i.e. free)
and available at http://www.rstudio.com/

## Installation

If you don't have RStudio installed on your computer:

- Go to the [RStudio download page](https://rstudio.com/products/rstudio/download/#download).
- Under Installers select RStudio x.yy.zzz - Windows XP/Vista/7/8 (where x, y, and z represent version numbers)
- Double click the file to install it
- Once it’s installed, open RStudio to make sure it works and you don’t get any error messages

If you already have R already install:

- Open RStudio, and click on “Help” > “Check for updates”. If a new version is available, quit RStudio, and download the latest version for RStudio

## Environment

RStudio combines (*File>New File>R Script*):

- A **Script editor** window where you can edit and send you commands to the console

- A **Console** is where you can type commands and see the output

- A **Workspace** shows all the active objects. The history tab shows a
list of commands used so far.

- The **Files** tab shows all the files and folder in your default workspace. The **Plots** tab will show all your graphs. The **Packages** tab will list a series of packages or add-ons needed to run certain processes.
For additional info see the **Help** tab.

Among the many advantages of RStudio, you will appreciate the user-adjustable interface (see *Options>Appearance>Layout*), the auto-synthax, the suggestion `function {package}` together a brief description of its use.

Again using our `fish` data set, we can write a script to import the data set as an object `fish` and create a very simple plot to visualize difference of reef fish richness by country.

```{r,  eval=T}
# import data set and create an objct in R studio + simple plot 
fish<-read.table('Data/reef_fish.txt', header=T, sep='\t', dec='.')
barplot(fish$richness, main="Top 10 reef fish Richness (Allen, 2000)", horiz=TRUE, names.arg=fish$country, cex.names=0.5, las=1)
```

**Note**: using the options in R studio, you can easily export the resulting plot as and image, a pdf, or to copy it to your clipboard.  




# R markdown

Our course is a set of R markdown files that have been rendered to html files and are hosted on a repository https://github.com/vianneydenis/attentive-suricate.git. 

Each Topic of the class has a corresponding R markdown file named Topic_X_topic.Rmd, where X is the topic number. Today for example the file Topic_1_Intro.Rmd allows you to play and review the class today. You can render today .Rmd by 'kniting' it in RStudio. Options allow you to 'Preview in Window' or to 'Preview in viewer Panel'

This repository working directory available on GitHub is the working directory of the attentive-suricate.Rproj file and it contains all of the .Rmd, .html and data files used to review this course. While you can can individually download and render the .Rmd files. It is better to 'fork' this repository to your GitHub account, then clone a copy of this forked repository  (from your own Github account to your computer). This only needs to be done once. 

## GitHub

Create an account on [GitHub](https://github.com/). You can **create your first repository** to later connect it into  RStudio.

## Git

Install Git on your computer

### Windows

1. Download the git installer for Windows [here](https://git-scm.com/). This will install bash, which is a command-line shell program for interacting with files and programs on your computer.

2. Run the installer that you just downloaded. Click “Next” as needed and when you get to the following options accept default.

3. That’s it! You should now have three new programs installed in a folder called Git under your Program Files.

### macOS

Follow the instructions below or view this [video tutorial](https://www.youtube.com/watch?v=9LQhwETCdwY&ab_channel=SarahStevens).

1. Go to this [list](https://sourceforge.net/projects/git-osx-installer/files/) and download an installer for Git.
   + If you are using OS X 10.9 and higher, choose the most recent version for “mavericks”.
   + If you are using an earlier version of OS X, choose the most recent version for “snow leopard”.

2. Run the installer you downloaded.

3. After installing Git, there will not be anything in your /Applications folder, because Git is a command line program.


**Note**: if you have OSX El Capitan you may experience some errors when running Git if you don’t have an updated XCODE.[Here](https://stackoverflow.com/questions/32893412/command-line-tools-not-working-os-x-el-capitan-sierra-high-sierra-mojave) is a description of the issue on StackOverflow and the solution is to open a Terminal window and type `xcode-select--install`.

### Linux 

You may already have Git installed. If not, you can try to install it via your dsitribution package manager. For Debian/Ubuntu run 'sudo apt-get install git' and for Fedora run 'sudo yum install git'.


## RStudio

File > New Project > Version Control 'Checkout a project from a version control repository'. Enter URL of your repository, (its name), and where you want to copy it on your computer. Once you validate:

+ you will create a .RProject. Any change in your repository will appear in 'Git'
+ changes needs to be 'commit' (prepared) and 'push' (uploaded) to GitHub in order to be synchronized
+ first 'commit' will fail and ask you to set up Git with email and username. Open Git CMD and process as indicated
+ first 'push' will be pending your GitHub login and password
+ any changes in your GitHub repo must be 'pull' prior working on your project in RStudio

## Rmarkdown

R markdown is text-based formatting that allows you to embed code and explanatory text in the same document.R markdown documents consist of a header, code chunks and text. R markdown files (.Rmd) can be rendered to other document formats (e.g. html, pdf, docx) to generate reports or web applications. For example, the present document was generated as an html from a Rmd file. Formatting analyses in R markdown keeps the entire research process in one document that can easily regenerate reports and output if the data are changed.

Visit R markdown [webpage](https://rmarkdown.rstudio.com/) for an overview of its capability. [Download](https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf) cheat sheet  to get familiar with it. I also recommend you R markdown bible: Xie et al. (2020), available [here](https://bookdown.org/yihui/rmarkdown/) to become an expert with markdown. 


As for now, just try in RStudio: 

*File > New File > R markdown* in order to create a simple .html document. Try to modify and play with it with the information available in the cheat sheets you downloaded.  

> Attribution
>
> The content of this webpage was derived from Fukami T. & Coyle J. (Standford University) which themselves derived from Data Carpentry and Software Carpentry materials (Copyright (c) Data Carpentry) under CC-3.0 Attribution License (see http://creativecommons.org/licenses/by/3.0/).





Next week: install and load a package, set up your working directory, load a dataset and look for some help on the application of a given function




