---
title: "dplyr"
---

```{r eval = TRUE,  message=F, include= FALSE, warning=F, purl=F, results="hide"}
knitr::purl('dplyr.Rmd', documentation = 0)
```


```{r echo=FALSE, purl=F}
xfun::embed_file('dplyr.Rmd')
```

```{r echo=FALSE, purl=F}
xfun::embed_file('dplyr.R')
```

The package `dplyr` is  specifically dedicated to the data manipulation in R. It has many common functions to fasten  data exploration and transformation. Some important functions from this packages are covered below.

```{r class.source = "fold-show",  eval=T}
library (dplyr)
```



# Summary

The function `summarise` is used to find insights (mean, median, mode, etc.) from a dataset. It reduces multiple values down to a single value. For example:

```{r,  eval=T}
summarised <- summarise(iris, Mean.Width = mean(Sepal.Width))
head(summarised)
```

# Manipulation

## `select`

It is used to select data by column name. Several syntaxes are possible. For example:

```{r class.source = "fold-show",  eval=T}
# by column names
selected <- dplyr::select(iris, Sepal.Length, Sepal.Width, Petal.Length)
head(selected) 

# by column range
selected1 <- dplyr::select(iris, Sepal.Length:Petal.Length)
head(selected1, 4)

# by column range number
selected2 <- dplyr::select(iris,c(2:5))
head(selected2)
```

Use `-` to hide a particular column

```{r,  eval=F}
selected <- select(iris, -Sepal.Length, -Sepal.Width)
```

## `filter`

It is used to filter rows with matching criteria. It works in teh same way than the function `select` (i.e., we pass a data frame along with a condition separated by a comma).

- Single condition:

```{r,  eval=T}
# Select setosa species
filtered <- filter(iris, Species == "setosa" )
head(filtered,3)
```

- Dual conditions:

```{r,  eval=T}
# Select versicolor species with Sepal width more than 3
filtered1 <- filter(iris, Species == "versicolor", Sepal.Width > 3)
tail(filtered1)
```


## `mutate`

It creates new columns (variables) and preserves the existing columns in a data set. For example:

```{r,  eval=T}
#To create a column “Greater.Half” which stores TRUE if given condition is TRUE
mutated1 <- mutate(iris, Greater.Half = Sepal.Width > 0.5 * Sepal.Length)
tail(mutated1)
```

To check how many flower satisfy this condition, the output of the function `table{base}` is a contingency table with the no. of individual (`FALSE` and `TRUE`):

```{r,  eval=T}
table(mutated1$Greater.Half)
```

## `arrange`

It is used to sort rows by variables in both an ascending and descending order. For example:

```{r,  eval=T}
# Sepal Width by ascending order
arranged <- arrange(iris, Sepal.Width)
head(arranged)

# Sepal Width by descending order
arranged1 <- arrange(iris, desc(Sepal.Width))
head(arranged1)
```

## `group_by`

It is done to group observations within a dataset by one or more variables. Data operations are often performed on groups defined by variables. For example when combined:

```{r,  eval=T}
# Mean sepal width by Species
gp <- group_by(iris, Species)
gp.mean <- summarise(gp,Mean.Sepal = mean(Sepal.Width))
gp.mean
```

Visit the follwoing [website](https://dplyr.tidyverse.org/) for more information on the `dplyr` package

# Pipe operator

The pipe operator `%>%` allows to wrap multiple functions together. It can be used with functions like `filter`, `select`, `arrange`, `summarise`, `group_by`, etc. As for  an example:

```{r,  eval=T}
#To select the rows with conditions
iris %>% filter(Species == "setosa",Sepal.Width > 3.8)
```

To find mean `Sepal`Length by `Species`, we use teh pipe operator as following:

```{r,  eval=T}
iris  %>% group_by(Species) %>% summarise(Mean.Length = mean(Sepal.Length))
```

